Is the direction of Granger causal influence the same as the direction of information flow?

### 2015-07-10 paper.tex
Add more acknowledgements.
Make Pulkit's edits.

\begin{abstract}
	Granger causality is an established measure of the ``causal influence'' that one stochastic process has on another. Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer causal influences. Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''. We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow? We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design. The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed. We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information. Thus, while it might be reasonable to infer the direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
\end{abstract}

@@

In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.). In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (eg.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (eg.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}). For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''. This raises the question: do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?

@@

In order to determine the direction of greater causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is \emph{compared} to that from $\{Y\}$ to $\{X\}$. The direction of causal influence is taken to be the direction with a greater Granger Causality metric. Further, this direction of causal influence is interpreted to be the direction of information flow -- the interpretation we question in this paper. We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency. It then becomes possible to also determine the brain wave frequency at which these influences occur. However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient to for the purpose of our arguments.

We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this is commonly done in practice even in areas beyond neuroscience (eg.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).

While we accept that Granger Causal influence can be accurately computed (provided measurements are taken suitably; see Section~\ref{sec:previous-criticisms-of-gc}) and can be useful in many situations (see Section~\ref{sec:conclusions}), interpreting the direction of Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.

@@

Several objections have been leveled against Granger Causality in the past. We give, here, a short overview of these and describe why our objection is novel, and possibly more fundamental in nature, at least in the context of its usage in neuroscience.

@@

@@

\begin{enumerate}
	\item Firstly, and most importantly, previous work has already shown that presence of a Granger Causal influence does not imply true causation. Judea Pearl classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model~\cite{pearl2009causality-gc-stat}. A potential issue, therefore, might be the question the novelty of this work, given that Pearl's classification could be seen to clear up the misconception. While our argument is, in spirit, very similar to this criticism of Granger causality, in the situation we study, we ask a different question - we wish to understand whether or not Granger Causality gives the correct direction of \emph{information flow}. Further, we give a concrete counter-example in this context, which is very relevant in the modern neuroscience setting.
	\item Our analysis does not tackle an information source that evolves with time. Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source.
	\item Our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work.
	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a trivial extension.
	\item We have also restricted ourselves to analyzing linear feedback communication strategies. In the presence of noise in the feedback link, linear communication strategies are known to be sub-optimal~\cite{YoungHanKimPaper}. In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal communication strategy -- which would have to be non-linear. This could be scope for future work. However, we do not expect results to change dramatically -- when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
	\item We consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
\end{enumerate}

@@

The authors would like to thank Nicola Elia for discussions that suggested this possible misunderstanding in use of Granger causality and Directed Information in computational systems. Praveen Venkatesh is partially supported by a dean's fellowship. %Pulkit Grover is supported by [...].

>>>>

\begin{abstract}
	Granger causality is an established statistical measure of the ``causal influence'' that one stochastic process has on another. Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer statistical causal influences. Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''. We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow? We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design. The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed. We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information. Thus, while it might be reasonable to infer the direction of statistical causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
\end{abstract}

@@

In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.). In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (e.g.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (e.g.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}). For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''. This raises the question: do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?

@@

In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$. The direction of causal influence is then usually taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}). Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper. We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency. It then becomes possible to also determine the brain wave frequency at which these influences occur. However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient to for the purpose of our arguments.

We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison. However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).

While we accept that it might be possible to accurately estimate Granger Causal influence (provided measurements are taken suitably; see Section~\ref{sec:previous-criticisms-of-gc}) and that it could be useful in many situations (see Section~\ref{sec:conclusions}), interpreting the direction of greater Granger Causal influence as the direction of information flow can be erroneous, as we demonstrate in this paper.

@@

Several objections have been raised against the use of Granger Causality in the past. We give, here, a short overview of these and describe why our objection is novel, and possibly more fundamental in nature, at least in the context of its usage in neuroscience.

@@

A more serious objection has to do with the difference between statistical measures of causality (such as Granger Causality) and true causality, and whether or not our work simply alludes to this difference. Our principal argument is different, however, as we describe in remark~\ref{remark:gc-vs-true} in Section~\ref{sec:objections-and-shortcomings}.

@@

\begin{enumerate}
	\item \label{remark:gc-vs-true} Previous work has already noted that Granger Causal influence does not imply true causation~\cite{granger1980testing}. This differentiation is made rigorously by Judea Pearl~\cite{pearl2009causality-gc-stat}, where he classifies Granger Causality as stemming from a ``statistical'' model, rather than from a ``causal'' model. The argument we make is very similar in spirit: we ask whether or not Granger Causality gives the correct direction of \emph{information flow}. A question on the novelty of our work may therefore be raised: if our argument boils down to a restatement of Pearl's differentiation, then this work has no new conceptual contribution. \\
		We make the case that our argument \emph{is} novel in the following manner: in the systems we consider -- the brain, as well as the communication system in our counter-example -- causal influences exist in both directions. In our counter-example, for instance, the feedback communication algorithm that is employed involves transmissions from both $Tx$ and $Rx$. The transmissions of each depend on what was transmitted by the other in the previous time instant. Causal influence and true causation, therefore, exist in both directions. The information of interest, however, flows in only one direction: from the transmitter to the receiver. We ask whether or not the direction of \emph{this} information flow can be discerned by comparing Granger causal influences in each direction. To this end, we give a concrete counter-example. This work is particularly relevant in the context of modern neuroscience, where such directions of information flow are desired in order to understand brain function.
	\item Our analysis does not tackle an information source that evolves with time. Hence, our communication process (inspired from the scheme of Schalkwijk and Kailath~\cite{S&K}) is non-ergodic. Since Granger causality is really just relative errors in prediction of a process, in presence or absence of knowledge of another process, we compute the obvious generalization of Granger causality to non-ergodic processes. Nevertheless, future work will address a situation with a linear dynamical system as the information source.
	\item Our experiments restrict themselves to Gaussian noise for simplicity, but neural spiking and spike-rate models for spikes tend to be very different from those used here. This is a clear direction for future work.
	\item The power and energy constraints are somewhat oversimplified to make the analysis simpler. This is for simplicity of exposition. A more general analysis is a simple extension.
	\item We have also restricted ourselves to analyzing linear feedback communication strategies. In the presence of noise in the feedback link, linear communication strategies are known to be sub-optimal~\cite{YoungHanKimPaper}. In order to make a water-tight argument, we would need to show that Granger Causality fails to correctly predict the direction of information flow, even when an optimal (non-linear) communication strategy is employed. This could be scope for future work. However, we do not expect results to change dramatically: when the feedback link is impaired by noise of only very low variance, the noiseless case should make for a good approximation of the system, and results should degrade gracefully, if there is any degradation at all.
	\item We consider a simple point-to-point network. In general networks, this issue could be even more complex. However, since this issue shows up even for the simplest network, we feel that the problem will only be exacerbated when the network is large.
\end{enumerate}

@@

The authors would like to thank Nicola Elia for discussions that suggested this possible misunderstanding in use of Granger causality and Directed Information in computational systems. We also thank Rob Kass, Momin Malik and Kun Zhang for useful discussions. Finally, we thank the reviewers of a previous submission of this work to ISIT for their insightful remarks and comments. Praveen Venkatesh is partially supported by a dean's fellowship. %Pulkit Grover is supported by [...].
